---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ExactVaRTest

Fast exact finite-sample back-testing for Value-at-Risk (VaR) models in R.

<!-- badges: start --><!-- badges: end -->

**ExactVaRTest** implements a forward dynamic programming algorithm that gives the exact finite-sample distributions and p-values of Christoffersen’s (1998) independence (IND) and conditional-coverage (CC) tests for Value-at-Risk (VaR) exception series. For completeness, it also provides the exact unconditional‑coverage (UC) test via a closed‑form binomial enumeration. 

In particular, it corrects the severe size distortions from which the usual asymptotic \(\chi^2\) approximation suffers in small samples and under extreme coverage rates.

* `backtest_lr()` returns the LR statistic, its exact p‑value, and a reject / fail‑to‑reject decision for one chosen test type (UC, IND, or CC).  

* `backtest_all()` runs the UC, IND, and CC tests jointly and returns summaries for all three statistics.

## Installation

You can install the development version of ExactVaRTest from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("YujianCHEN219/ExactVaRTest")
```

## Example

```{r example}
library(ExactVaRTest)

set.seed(42)
x <- rbinom(300, 1, 0.03)          # synthetic 0/1 exception series

bt <- backtest_lr(x, alpha = 0.05, type = "cc")  # exact LR_cc back-test
print(bt)
```

## Main features

* Exact finite‑sample distributions and p‑values for LR\_ind and LR\_cc at any sample size *n*; for *n* ≤ 2000 the computation finishes in milliseconds to a few seconds.  

* C++ implementation via `Rcpp`, with automatic fallback to a pure‑R reference engine.  

* Minimal dependencies (`Rcpp`, `stats`); works on macOS, Linux, and Windows.

## Extensions

Freely extends to *CoVaR backtesting*: pass the institution’s hit sequence on system‑VaR‑breach days into `backtest_lr()` for exact UC/IND p‑values; for short windows or extreme tails, one option is to treat the systemic‑breach count as random and apply the mixture‑tail test to maintain correct size. (see vignettes and [Francq & Zakoïan 2025]).

## Suggested readings 

1. Christoffersen, P. F. (1998). *Evaluating interval forecasts.* International economic review, 841-862.

2. Mehta, C. R., Patel, N. R., & Gray, R. (1985). *Computing an exact confidence interval for the common odds ratio in several 2× 2 contingency tables.* Journal of the American Statistical Association, 80(392), 969-973.

3. Francq, C., & Zakoïan, J. M. (2025). Inference on dynamic systemic risk measures. Journal of Econometrics, 247, 105936.

## Acknowledgements

I greatly appreciate Christian Francq, Christophe Hurlin, and Jean-Michel Zakoïan's guidance and support.

In particular, Christian Francq generously shared the initial idea; without his help, this package would not exist.

## License

This package is free and open source, licensed under GPL.
